Kernel k-Groups via Hartigan's Method

### 2018-08-10 energy_clustering/pami18/energy_clustering.tex
minor updates

\title{Kernel k-Groups and Hartigan's Method}

\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua~Vogelstein
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem G. Fran\c ca 
and J. Vogelstein are with the Center for Imaging Science, Johns Hopkins
University; J. Vogelstein is also with 
the Department of Biomedical Engineering and Institute for Computational
Medicine, Johns Hopkins University. \protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: guifranca@jhu.edu; jovo@jhu.edu
\IEEEcompsocthanksitem M. Rizzo is with the Department of Mathematics
and Statistics, Bowling Green State University. \protect\\
E-mail: mrizzo@bgsu.edu}% <-this % stops an unwanted space
\thanks{Manuscript received March 15, 2018; revised March 15, 2018.}}

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
{Energy Clustering}

@@

Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering 
by extending Ward's method of minimum variance
\cite{RizzoClustering}; 
see \cite{Szkely2013,Szkely2017} for an overview of energy
statistics and its applications.
Moreover, in Euclidean spaces, an application of 
energy statistics to clustering
was recently proposed \cite{Kgroups}, and the method was named 
\emph{k-groups}.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
More recently, the notion of distance covariance was further 
generalized from Euclidean 
spaces to metric spaces of negative type \cite{Lyons}. Furthermore, 
the link between energy distance based tests and kernel 
based tests has 
been recently established \cite{Sejdinovic2013} 
through an equivalence between generalized energy distances and maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS).
This equivalence immediately relates energy statistics to
kernel methods often used in machine learning, and form the basis 
of our approach in this paper.

@@

\subsection*{Contributions}
Although k-groups considers clustering from energy statistics
in the particular Euclidean case \cite{Kgroups},
the precise optimization problem behind this approach
remains obscure, as well as the connection with other methods
in machine learning.
The main theoretical contribution of this paper is to fill these gaps, 
which 
we do so in more generality. For instance, our approach is not limited
to the Euclidean case but holds for general arbitrary spaces
of negative type. 
For generality, we also
consider a weighted version of energy statistics.
Our approach reveals connections
between energy statistics based clustering and existing methods
such as kernel k-means and graph partitioning problems.

@@

The optimization problem \eqref{eq:qcqp3} is nonconvex, besides
being NP-hard, thus a direct approach 
is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints. For instance, consider the relaxed problem
\begin{equation}
\label{eq:relaxed}
\max_{H} \Tr \left[ H^\top \widetilde{G} \, H \right] 
\qquad \mbox{such that $H^\top H = I$},
\end{equation}
where $\widetilde{G} \equiv \mathcal{W}^{1/2} \, G \, \mathcal{W}^{1/2}$.
This problem has a well-known closed form solution $H^\star = U R$, where the
columns of $U \in \mathbb{R}^{n\times k}$ 
contain the top $k$ eigenvectors of $\widetilde{G}$ corresponding
to the $k$ largest eigenvalues, $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and $R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting optimal objective function assumes the value 
$\max \Tr \big[ {H^\star}^\top \widetilde{G} \, H^\star \big]  = 
\sum_{i=1}^k \lambda_i$. 
Spectral clustering is based on this approach, where
one further normalize the rows of $H^\star$, then cluster
the resulting rows as data points using any clustering method such
as k-means.
A procedure on these lines was proposed in the seminal papers
\cite{Malik,NgJordan}.

>>>>

\title{Kernel k-Groups via Hartigan's Method}

\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua T.~Vogelstein
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem G. Fran\c ca 
and J. Vogelstein are with the Center for Imaging Science, Johns Hopkins
University; J. Vogelstein is also with 
the Department of Biomedical Engineering and Institute for Computational
Medicine, Johns Hopkins University. \protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: guifranca@jhu.edu; jovo@jhu.edu
\IEEEcompsocthanksitem M. Rizzo is with the Department of Mathematics
and Statistics, Bowling Green State University. \protect\\
E-mail: mrizzo@bgsu.edu}% <-this % stops an unwanted space
\thanks{Manuscript received March 15, 2018; revised March 15, 2018.}}

% The paper headers
\markboth{Energy Clustering}{}
%{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%

@@

Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation~\cite{Szekely2007-mm}, which generalizes the Pearson
correlation coefficient, and hierarchical clustering 
by extending Ward's method of minimum variance
\cite{RizzoClustering}; 
see \cite{Szkely2013,Szkely2017} for an overview of energy
statistics and its applications.
Moreover, in Euclidean spaces, an application of 
energy statistics to clustering
was recently proposed \cite{Kgroups}, and the method was named 
\emph{k-groups}.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
More recently, the notion of distance covariance was further 
generalized from Euclidean 
spaces to metric spaces of negative type \cite{Lyons}. Furthermore, 
the link between energy distance based tests and kernel 
based tests has 
been recently established \cite{Sejdinovic2013} 
through an asymptotic equivalence between generalized energy distances and maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS).
Even more recently, generalized energy distances and kernel methods have been demonstrated to be exactly equivalent, for all finite samples~\cite{Shen2018-st}.
This equivalence immediately relates energy statistics to
kernel methods often used in machine learning, and form the basis 
of our approach in this paper.

@@

\subsection*{Contributions}
Although k-groups considers clustering from energy statistics
in the particular Euclidean case \cite{Kgroups},
the precise optimization problem behind this approach
remains obscure, as well as the connection with other methods
in machine learning.
The main theoretical contribution of this paper is to fill these gaps, 
which 
we do   more generality. For instance, our approach is not limited
to the Euclidean case but holds for general arbitrary spaces
of negative type. 
For generality, we also
consider a weighted version of energy statistics.
Our approach reveals connections
between energy statistics based clustering and existing methods
such as kernel k-means and graph partitioning problems.

@@

The optimization problem \eqref{eq:qcqp3} is nonconvex, besides
being NP-hard, thus a direct approach 
is computationally prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints. For instance, consider the relaxed problem
\begin{equation}
\label{eq:relaxed}
\max_{H} \Tr \left[ H^\top \widetilde{G} \, H \right] 
\qquad \mbox{such that $H^\top H = I$},
\end{equation}
where $\widetilde{G} \equiv \mathcal{W}^{1/2} \, G \, \mathcal{W}^{1/2}$.
This problem has a well-known closed form solution $H^\star = U R$, where the
columns of $U \in \mathbb{R}^{n\times k}$ 
contain the top $k$ eigenvectors of $\widetilde{G}$ corresponding
to the $k$ largest eigenvalues, $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and $R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting optimal objective function assumes the value 
$\max \Tr \big[ {H^\star}^\top \widetilde{G} \, H^\star \big]  = 
\sum_{i=1}^k \lambda_i$. 
Spectral clustering is based on this approach, where
one further normalize the rows of $H^\star$, then cluster
the resulting rows as data points using any clustering method such
as k-means.
A procedure on these lines was proposed in the seminal papers
\cite{Malik,NgJordan}.
